---
title: 'Homework: Finite Precision Arithmetic & Gradient Descent'
output:
  html_document:
    df_print: paged
  html_notebook:
    code_folding: none
    highlight: textmate
---


```{r setup, include=FALSE}
rm(list = ls())
source(file.path("..", "R", "colors.R"))
source(file.path("..", "R", "util.R"))

required_packages <- c('Matrix', 'testthat')
install_and_load_packages(required_packages)
```


# Exercise 1: One-pass algorithm for calculating variance

Let's say you are running a Monte Carlo simulation to estimate the mean and variance of some random variable $\boldsymbol{\theta}$. 
Imagine that $\boldsymbol{\theta}$ is high-dimensional, so you don't want to keep all the Monte Carlo samples $\boldsymbol{\theta}_i$ due to the memory burden.
(If $\boldsymbol{\theta}$ is $10^6$-dimensional, for example, storing 1,000 Monte Carlo samples would require 8GB of memory.)
And it occurs to you that you can write a "one-pass" algorithm to calculate the sample mean and sample second moment of $\boldsymbol{\theta}$ by only keeping track of running totals of Monte Carlo samples. 
You can then, in theory, calculate the sample variance from these quantities.

The above idea is implemented and applied below. 
Try it and see what you get. 
Does the estimate make sense? If not, why?
How does the estimate change if you change the seed?


**Bonus question:** 
What is, roughly speaking, the distribution of the estimator `sample_var_with_question_mark` under repeated Monte Carlo simulations (i.e. without a fixed seed)?

```{r}
set.seed(140778)

running_sum <- 0
running_sum_sq <- 0

n_sample <- 10^3
for (i in 1:n_sample) {
  theta <- 1e10 + rnorm(1)
  running_sum <- running_sum + theta
  running_sum_sq <- running_sum_sq + theta^2
}
sample_mean <- running_sum / n_sample
sample_sec_mom <- running_sum_sq / n_sample

sample_var_with_question_mark <- (sample_sec_mom - (sample_mean)^2)
```

## Things to commit/submit
Answers to the questions above.

**Answer**
The estimate does not make sense since the variance can never be negative.
Reasons: In R, the numbers are stored in the 64-bits double-precision floating-point format.
If the true sum of the two numbers contains more than 64 bits, then the excess bits will be lost when the resultis rounded to 64 bits.
Here, when we calculate theta, we summing up two numbers, 1e10 and rnorm(1), where there's great difference in magnitude.
Therefore, the entire mantissa of the smaller number may be shifted beyond the field width so that thetamay not exactly match the true summation of the two numbers.
The same will happen when we get the product of the numbers, a.k.a, when we calculating $\theta^2$.
The product of two p-digit mantissas will in general contain up to 2p digits, and thus once the correct result cannot be represented exactly in the floating-point system and must be rounded.

When the seed is 140778, the sample_var_with_question_mark has a negative value of -147456.
when the seed is 42, the sample_var_with_question_mark has a value of 114688, which still does not make sense since the true variance is 1.


# Exercise 2: Numerical differentiation

## Part A: Potential caveats of numerical differentiation

You have learned in the class that the centered difference $\frac{f(x + \Delta x) - f(x - \Delta x)}{2 \Delta x}$ approximates the derivative $f'(x)$ up to an $O(\Delta x^2)$ error.
To numerically verify this, we design the following experiment using a log of logistic function (i.e. $f(x) = \log\{ \exp(x) / (1 + \exp(x)) \}$) as an example.
We evaluate its derivative using an analytical formula and compare it to the centered difference approximations as $\Delta x$ varies from $2^{-1}$ to $2^{-52}$.
Using the starter code provided below, plot the relative error of the numerical approximation in a log-log scale.
What do you expect the slope of the plotted line to be?


```{r, eval=TRUE}
#' Approximate derivative via centered-difference approximation
approximate_deriv <- function(func, x, dx) {
  return ((func(x+dx)-func(x-dx))/2/dx)
}

log_logistic_func <- function(x) {
  # logistic(x) = exp(x)/(1+exp(x))
  # log-logistic(x) = x - log(1+exp(x))
  return (x - log(1+exp(x)))
}

log_logistic_deriv <- function(x) {
# log-transformation of the log-logistic function
  return (1/(1+exp(x)))
}

# Calculate and plot the relative errors of finite difference approximations
set.seed(615)
x <- rnorm(1)

log2_dx <- - seq(1, 52)
numerical_deriv <- sapply(
  2^log2_dx, function(dx) approximate_deriv(log_logistic_func, x, dx)
)
analytical_deriv <- log_logistic_deriv(x)
rel_err <- abs(
  (analytical_deriv - numerical_deriv) / analytical_deriv
)

fontsize <- 1.2
plot(
  log2_dx, log2(rel_err),
  frame = FALSE, # Remove the ugly box
  col = jhu_color$spiritBlue,
  xlab = "log2 stepsize for finite difference",
  ylab = "log2 rel errors",
  xlim = rev(range(log2_dx)),
  cex.lab = fontsize,
  cex.axis = fontsize,
  type = 'l'
)
```

**Questions:**
Does the relative error plot agree with what the mathematical theory predicts? Explain the observed phenomenon.

**Answer**
The relative error plot does not agree with the mathematical theory. In theory, we expect the line to have a negative slope. 
As $\Delta x$ decreases (much smaller than 1), the scale of the error will decrease given $O(\Delta x^2)$ will be small and negligible.
However, here we saw the relative error goes down first as $\Delta x$ gets smaller; while after $\Delta x$ gets to $2^-{18}$, the relative error starts going up. 

**Questions:**
Also, what is the slope in the "latter part" of your plotted line, roughly speaking? 
Explain why we observe such a slope in the relative error. 
(Hint: Consider a machine with 10 decimal digits of accuracy. What happens if we numerically differentiate $f(x) = x$ at $x = 1$ with $\Delta x = 0.1, 0.01, 0.001, \ldots$.)

**Answer**
The slope in the latter part of our plotted line is around 1.

For this question, the relative error is $\frac{O(\Delta x^2)+\epsilon}{2\Delta xf'(x)}$, where $\epsilon$ is the machine epsilon.
Before the tipping point, the relative error is dominated by $O(\Delta x^2)/f'(x)$, therefore, the slope was roughly -2. This happens when $\Delta x$ is not too small so that $f(x+\Delta x)-f(x-\Delta x)$ can be computed in finite-precision with reasonable error.
However, when $\Delta x$ gets smaller, the relative error is dominated by the machine epsilon, so that the slope is roughly $-\frac{log_2(\epsilon/(2\Delta x_1f'(x)))-log_2(\epsilon/(2\Delta x_2f'(x)))}{log_2(\Delta x_1)-log_2(\Delta x_2)}=1$.

## Part B: Extention to multivariate functions

As mentioned in the lecture, the finite difference method extends to a function with multivariate input $f: \mathbb{R}^d \to \mathbb{R}$ and can be used to approximate the gradient $\nabla f(\boldsymbol{x})$.
The extension relies on the fact $f(\boldsymbol{x} + \Delta x \, \boldsymbol{e}_i) = f(\boldsymbol{x}) + \Delta x \, \partial_i f(\boldsymbol{x}) + \ldots$ for $i = 1, \ldots, d$.

Using the starter code provided below, implement a function to approximate $\nabla f(\boldsymbol{x})$ via the centered difference method.
To test this function, take a quadratic function $f(x) = - \boldsymbol{x}^\intercal \boldsymbol{\Phi} \boldsymbol{x} / 2$ with positive definite $\boldsymbol{\Phi}$ as an example.
First derive an analytical expression for its gradient.
Then compare a numerically approximated gradient to one computed via an analytical formula.

(In deriving an analytical formula for the gradient, you might find these hints helpful: 
1) any positive definite $\boldsymbol{\Phi}$ admits a "square-root" decomposition $\boldsymbol{\Phi} = \boldsymbol{A}^\intercal \boldsymbol{A}$ and hence $\boldsymbol{x}^\intercal \boldsymbol{\Phi} \boldsymbol{x} = \| \boldsymbol{A} \boldsymbol{x} \|^2$; and
2) by the chain rule, the gradient of $g(\boldsymbol{y})$ for $\boldsymbol{y} = \boldsymbol{A} \boldsymbol{x}$ with respect to $\boldsymbol{x}$ is $\nabla_{\boldsymbol{x}} \left( g \circ \boldsymbol{A} \right) (\boldsymbol{x}) = \boldsymbol{A}^\intercal \nabla_{\boldsymbol{y}} g(\boldsymbol{y})$.)

```{r, eval=FALSE}
are_all_close <- function(v, w, abs_tol = 1e-6, rel_tol = 1e-6) {
  abs_diff <- abs(v - w)
  are_all_within_atol <- all(abs_diff < abs_tol)
  are_all_within_rtol <- all(abs_diff < rel_tol * pmax(abs(v), abs(w)))
  return(are_all_within_atol && are_all_within_rtol)
}

# approx_grad_for_loop <- function(func, x, dx = .Machine$double.eps^(1/3)) {
#   # finite difference for multivariate
#   d <- length(x)
#   numerical_grad <- rep(0, d)
#   for (i in 1:d){
#     delta_x <- rep(0, d)
#     delta_x[i] <- dx
#     numerical_grad[i] <- (func(x + delta_x) - func(x - delta_x)) / (2*dx)
#   }
#   return(numerical_grad)
# }

approx_grad <- function(func, x, dx = .Machine$double.eps^(1/3)) {
  # finite difference for multivariate
  return (sapply(1:length(x), function(i) approx_partial_grad(i, func, x, dx)))
}

approx_partial_grad <- function(i, func, x, dx = .Machine$double.eps^(1/3)) {
  delta_x <- rep(0, length(x))
  delta_x[i] <- dx
  return ((func(x + delta_x) - func(x - delta_x)) / (2*dx))
}

set.seed(410)
n_param <- 4
X <- matrix(rnorm(2 * n_param^2), nrow = 2 * n_param, ncol = n_param)
SIGMAINV <- t(X) %*% X

#' Calculate log-density of centered Gaussian up to an additive factor
gaussian_logp <- function(x, Sigma_inv = SIGMAINV) {
  logp <- - .5 * t(x) %*% Sigma_inv %*% x
  return(logp)
}

gaussian_grad <- function(x, Sigma_inv  = SIGMAINV) {
  grad <- - Sigma_inv %*% x
  return(grad)
}

x <- c(3, 1, 4, 1)

analytical_grad <- gaussian_grad(x, SIGMAINV)
numerical_grad <- approx_grad(func = function(x) gaussian_logp(x, SIGMAINV), x = x)
testthat::expect_true(are_all_close(
  analytical_grad, numerical_grad, abs_tol = Inf, rel_tol = 1e-3
))
```


## Things to commit/submit
For Part A, completed code and answers to the questions.
For Part B, completed code that passes the test.


# Exercise 3: Continuing development of `hiperglm` &mdash; finding MLE via `stats::optim()` and testing it

We will continue the development of the `hiperglm` package. 

As discussed, our plan is to first implement an MLE finder via a pre-packaged general-purpose BFGS optimizer, against which we can then use to benchmark our custom implementation of iteratively reweighted least squares. 
The idea is that using the pre-packaged function is less error-prone, especially since the gradient function is easy to unit-test.

To be extra safe, we are going to test the MLE finder via BFGS against an even simpler algorithm: the least square for a linear model via the analytical expression $\hat{\boldsymbol{\beta}}_\textrm{mle} = (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal \boldsymbol{y}$.
Complete the following steps toward this goal.
Remember that you can use [`devtools::load_all()`](https://r-pkgs.org/code.html#sec-code-load-all) to help your [package development process](https://r-pkgs.org/code.html#constant-health-checks). 

0. Make and check out (`git branch` and `git checkout`) a development branch (call it `dev` or `develop`) where you will commit the work below.
1. Set up the use of `testthat`.
2. Copy the helper functions `are_all_close()` and `simulate_data()` from the instructor's repo.
3. Test the `are_all_close()` function. 
Write a test to cover at least three cases: the function 1) correctly returns `TRUE`, 2) correctly returns `FALSE` because the relative error is above `rel_tol`, and 3) correctly returns `FALSE` because the absolute error is above `abs_tol`. 
Make sure that the logic of the test is clear and easy to understand &mdash; tests are there to clarify and validate the code's behavior; unreadable tests defeat the purpose.
<br> _Remark:_ This is just an exercise, but in general there is nothing strange about testing a function you use to test another function. It is perfectly reasonable to consider testing any functions complicated enough to potentially contain insiduous bugs.
4. Set up the (currently failing) test to compare the MLE estimated via pseudo-inverse and via BFGS. 
5. Implement the MLE finder via peudo-inverse. 
Please, pretty please, do not calculate an inverse of a matrix. 
(Otherwise, it's fine to use `solve()` instead of `chol()`.)
6. Implement the MLE finder via BFGS:
    a. Implement functions to calculate the log-likelihood and gradient under a linear model. 
    This function technically requires an estimate of the residual variance, but its value does not affect the MLE for the coefficients and hence is irrelevant for our purpose. 
    You can just specify it as an optional parameter e.g. as `function(..., noise_var = 1)`.
    b. Test your gradient calculation by comparing it against a numerical one.
    For this purpose, use the function from Exercise 2, Part B to approximate the gradient of a given function via finite difference.
    <br> _Remark:_ Technically, this test checks only for consistency between the log-likelihood and gradient function.
    It is possible to have both functions wrong but consistent with each other so that the test passes.
    This test nonetheless can catch many bugs and, since the gradient calculation tends to be more error prone, it's good to make sure there is no mistake there. 
    c. Plug in the now tested-and-true log-likelihood and gradient functions to `stats::optim()` to find MLE.
7. Run `devtools::test()` and check that the then-failing test from Step 2 now passes.
8. Open a pull request from your development branch to the main/master one. 
    
## Things to commit/submit

Link to the pull request: https://github.com/graced03/hiperglm/pull/1
